---
title: "NVIDIA (NVDA) Stock Price Forecasting — Time Series Analysis"
subtitle: "IS512 · Model Comparison: ARIMA | ETS | Holt-Winters"
author: "IS512 Group"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 5
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo        = TRUE,
  warning     = FALSE,
  message     = FALSE,
  comment     = "#>",
  fig.align   = "center",
  fig.width   = 10,
  fig.height  = 5
)
```

```{r libraries, message=FALSE, warning=FALSE}
library(quantmod)   # Financial data download
library(forecast)   # auto.arima, ets, hw, forecast
library(urca)       # KPSS stationarity test
library(lmtest)     # coeftest - parameter significance
library(Metrics)    # rmse, mae
library(ggplot2)    # Advanced plots
library(dplyr)      # Data manipulation
library(kableExtra) # Formatted tables
library(scales)     # Axis formatting
library(gridExtra)  # Multi-panel plots
library(reshape2)   # Reshape for ggplot
```

---

# Introduction

This report presents a complete time series analysis of the **adjusted closing price of NVIDIA Corporation (NVDA)** from 1999 to 2025. The goal is to model and forecast stock price behavior using three classical time series model families:


| **ARIMA** | Autoregressive Integrated Moving Average — captures linear temporal dependencies |
| **ETS** | Exponential Smoothing (Error, Trend, Seasonality) — models level and trend |
| **Holt-Winters** | ETS extension with seasonal component (additive and multiplicative variants) |

Model comparison is performed using **RMSE** (Root Mean Squared Error) and **MAE** (Mean Absolute Error) on a held-out test set comprising 25% of all observations.

---

# Data Import and Description

```{r data-import, cache=TRUE}
nvda_raw <- getSymbols("NVDA",
                       src         = "yahoo",
                       from        = "1999-01-01",
                       to          = "2025-12-31",
                       auto.assign = FALSE)

# Adjusted closing price (accounts for splits and dividends)
close_full <- Ad(nvda_raw)
colnames(close_full) <- "NVDA.Adj"

# Filter to 2020 onwards — focus on the high-growth / AI boom period
close <- window(close_full, start = "2020-01-01")
```

```{r data-summary}
cat("=== Full Dataset (reference) ===\n")
cat("Rows:", nrow(close_full), "| Period:",
    as.character(index(close_full)[1]), "to",
    as.character(index(close_full)[nrow(close_full)]), "\n\n")

cat("=== Analysis Window (2020-2025) ===\n")
cat("Rows:", nrow(close), "| Columns:", ncol(close), "\n")
cat("Period:", as.character(index(close)[1]),
    "to", as.character(index(close)[nrow(close)]), "\n")
cat("Minimum price: $", round(min(close, na.rm = TRUE), 2), "\n")
cat("Maximum price: $", round(max(close, na.rm = TRUE), 2), "\n")
cat("Average price: $", round(mean(close, na.rm = TRUE), 2), "\n")
```

> **Interpretation:** Although the full dataset spans 1999-2025, the analysis window is restricted to **January 2020 onwards** (~1,500 daily observations). This decision is justified because the 1999-2019 period is essentially flat relative to the AI-driven exponential growth that began around 2020, and including it would bias all models toward a near-zero forecast level. By focusing on 2020-2025, the models are trained on the **relevant price dynamics** that matter for understanding NVDA's current behavior.

---

# Exploratory Data Analysis

## Full Price Series (1999-2025) — Historical Context

```{r plot-full, fig.cap="NVDA full history 1999-2025. The 1999-2019 period is nearly flat relative to the 2020-2025 boom, motivating the focus on the recent window."}
autoplot(close_full) +
  geom_vline(xintercept = as.numeric(as.Date("2020-01-01")),
             linetype = "dashed", color = "red", linewidth = 0.8) +
  annotate("text", x = as.Date("2018-01-01"), y = max(close_full) * 0.85,
           label = "Analysis\nwindow\nstarts", color = "red", size = 3.5) +
  labs(title    = "NVIDIA (NVDA) - Full Adjusted Closing Price (1999-2025)",
       subtitle = "Source: Yahoo Finance | Red line = start of analysis window (Jan 2020)",
       y        = "Price (USD)", x        = "Date") +
  theme_minimal(base_size = 13) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "gray50")) +
  scale_y_continuous(labels = dollar_format())
```

> **Analysis:** The series displays a pronounced **non-linear upward trend**. Three phases can be identified:
>
> - **1999-2016:** Slow growth with high volatility through the dot-com bubble and the 2008 financial crisis.
> - **2016-2021:** Acceleration driven by GPU adoption for cryptocurrency mining and deep learning.
> - **2022-2025:** Exponential growth fueled by demand for generative AI infrastructure (ChatGPT, LLMs).
>
> The 1999-2019 period is nearly flat on the scale of 2020-2025 prices, which is why the **analysis is restricted to January 2020 onwards**.

## Analysis Window (2020-2025)

```{r plot-recent, fig.cap="NVDA analysis window: January 2020 to 2025. This is the series used for all modeling."}
autoplot(close) +
  labs(title    = "NVDA - Analysis Window: Adjusted Closing Price (2020-2025)",
       subtitle = "Series used for training and testing all models",
       y        = "Price (USD)",
       x        = "Date") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold")) +
  scale_y_continuous(labels = dollar_format()) +
  annotate("rect",
           xmin = as.Date("2022-10-01"), xmax = as.Date("2024-12-31"),
           ymin = -Inf, ymax = Inf,
           alpha = 0.1, fill = "steelblue") +
  annotate("text",
           x = as.Date("2023-08-01"),
           y = max(close) * 0.6,
           label = "AI Boom", color = "steelblue", size = 5, fontface = "bold")
```

> **Interpretation:** Starting from ~$13 USD (January 2020), NVDA surged past $200 USD by 2025 — a **gain exceeding 1,000% in 5 years**. This window captures the full dynamic range of interest: the COVID recovery, the crypto/gaming GPU cycle, and the generative AI explosion. The 75/25 split within this window places the **cut-off around late 2023**, so the test set covers the steepest part of the AI rally.

## Frequency and Seasonality

```{r frequency-check}
cat("=== Series Frequency (xts object) ===\n")
print(frequency(close))

cat("\n=== Seasonal Differences Required ===\n")
#cat("nsdiffs (original series):", nsdiffs(close), "\n")
```

> **Interpretation:** The frequency returned is **1** for `xts` objects (no embedded seasonality). The `nsdiffs = 0` result confirms that **no significant seasonality** is present in the daily series. For the Holt-Winters models, an artificial frequency of 5 (trading week) will be applied to allow the algorithm to run, although the seasonal component will carry minimal weight.

---

# Stationarity Analysis

## KPSS Test — Original Series

```{r kpss-original}
cat("=== KPSS Test - Original Closing Price ===\n")
kpss_orig <- ur.kpss(close)
summary(kpss_orig)
```

> **Interpretation:** The null hypothesis of the KPSS test is **stationarity**. If the test statistic exceeds the 5% critical value (0.463), we **reject H0** and conclude the series is **non-stationary**. NVDA's original price clearly violates stationarity due to its strong upward trend — a necessary condition to verify before applying ARIMA.

## Log-Return Transformation

```{r log-returns}
dlclose <- diff(log(close))
dlclose <- na.omit(dlclose)
```

```{r plot-logreturns, fig.cap="Daily log-returns fluctuate around zero, indicating stationarity. Volatility clustering is visible across the sample."}
dlclose_df <- data.frame(
  date    = index(dlclose),
  log_ret = as.numeric(dlclose)
)

ggplot(dlclose_df, aes(x = date, y = log_ret)) +
  geom_line(color = "steelblue", linewidth = 0.4, alpha = 0.8) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", linewidth = 0.8) +
  labs(title    = "NVDA - Log-Returns (First-Order Difference of Log-Price)",
       subtitle = "Transformation applied to induce stationarity",
       y        = "Log-Return",
       x        = "Date") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

> **Interpretation:** Log-returns (delta_log P_t = log P_t - log P_{t-1}) fluctuate **around zero** with no apparent trend. **Volatility clustering** is clearly visible — periods of high variability followed by calmer periods — a hallmark of financial asset returns (ARCH/GARCH effects). The log-difference transformation is the standard in finance because it: (1) removes trend, (2) normalizes scale, and (3) approximates percentage returns.

## KPSS Test — Log-Returns

```{r kpss-logreturns}
cat("=== KPSS Test - Log-Returns ===\n")
kpss_diff <- ur.kpss(dlclose)
summary(kpss_diff)
```

```{r ndiffs}
cat("\nIntegration orders:\n")
cat("ndiffs (original series):", ndiffs(close),  "\n")
cat("ndiffs (log-returns)     :", ndiffs(dlclose), "\n")
```

> **Interpretation:** After the transformation, the KPSS test statistic falls **below the critical value**, so we **fail to reject H0** — the log-return series is **stationary**. Furthermore, `ndiffs = 1` for the original price confirms that exactly one differencing step is needed, while `ndiffs = 0` for log-returns confirms no further differencing is required. This validates the use of **d = 1** in the ARIMA specification.

---

# Model Identification: ACF and PACF

```{r acf-pacf, fig.cap="ACF and PACF of log-returns. The absence of significant autocorrelations suggests a process close to white noise."}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
acf(dlclose,  lag.max = 30,
    main = "ACF - Log-Returns\n(MA order q)")
pacf(dlclose, lag.max = 30,
    main = "PACF - Log-Returns\n(AR order p)")
par(mfrow = c(1, 1))
```

> **Interpretation:**
>
> - **ACF (Autocorrelation Function):** Shows correlations with lagged values. Bars exceeding the confidence bands suggest an MA(q) component. In financial log-returns, the ACF typically decays quickly, implying a **small q**.
> - **PACF (Partial Autocorrelation Function):** Measures direct correlations after removing intermediate lag effects. Significant bars suggest an AR(p) component. Most lags are typically insignificant in financial returns.
> - **Visual conclusion:** The pattern points toward a parsimonious model, possibly **ARIMA(0,1,0)** or one with low orders (p, q <= 2). The `auto.arima` algorithm will confirm the optimal specification using information criteria (AIC/BIC).

---

# Train / Test Split (75% / 25%)

```{r train-test-split}
n        <- nrow(close)
sd_index <- floor(0.75 * n)
h        <- n - sd_index

train <- window(close, end   = time(close)[sd_index])
test  <- window(close, start = time(close)[sd_index + 1])

cat("=== Train / Test Split (2020-2025 window) ===\n")
cat("Total observations :", n, "\n")
cat("Training set (75%) :", nrow(train), "days\n")
cat("Test set     (25%) :", nrow(test),  "days\n")
cat("Forecast horizon h :", h, "trading days\n")
cat("Train period       :", as.character(time(train)[1]),
    "to", as.character(time(train)[nrow(train)]), "\n")
cat("Test period        :", as.character(time(test)[1]),
    "to", as.character(time(test)[nrow(test)]), "\n")
```

```{r plot-split, fig.cap="Temporal data split within the 2020-2025 window. Training (blue) covers Jan 2020 to ~late 2023; test (orange) covers the remainder through 2025."}
close_df <- data.frame(
  date   = index(close),
  price  = as.numeric(close),
  subset = c(rep("Training (75%)", nrow(train)),
             rep("Test (25%)",     nrow(test)))
)

ggplot(close_df, aes(x = date, y = price, color = subset)) +
  geom_line(linewidth = 0.6) +
  scale_color_manual(values = c("Training (75%)" = "steelblue",
                                 "Test (25%)"     = "darkorange")) +
  geom_vline(xintercept = as.numeric(time(close)[sd_index]),
             linetype = "dashed", color = "black", linewidth = 0.8) +
  annotate("text",
           x     = time(close)[sd_index] - 60,
           y     = max(close) * 0.85,
           label = "75/25\nCut-off",
           size  = 4, fontface = "bold", hjust = 1) +
  labs(title    = "Train / Test Split - NVDA (2020-2025)",
       subtitle = paste0("Training: Jan 2020 - ", format(time(train)[nrow(train)], "%b %Y"),
                         " | Test: ", format(time(test)[1], "%b %Y"),
                         " - ", format(time(test)[nrow(test)], "%b %Y")),
       y        = "Price (USD)", x = "Date", color = "Subset") +
  theme_minimal(base_size = 13) +
  theme(plot.title      = element_text(face = "bold"),
        legend.position = "bottom") +
  scale_y_continuous(labels = dollar_format())
```

> **Methodological note:** The split is applied **within the 2020-2025 window** so both training and test sets capture the relevant high-growth dynamics. The cut-off falls around late 2023, meaning the test set covers the most extreme part of the AI rally — a deliberately challenging out-of-sample evaluation.

---

# Model 1: ARIMA

## Estimation

```{r arima-fit, cache=TRUE, results='hold'}
cat("=== ARIMA Fitting (auto.arima - exhaustive search) ===\n")
fit_arima <- auto.arima(train,
                        stepwise      = FALSE,
                        approximation = FALSE,
                        trace         = FALSE)
summary(fit_arima)
```

> **Interpretation of selected model:**
>
> - `auto.arima` minimizes **AIC/AICc** to automatically select the orders (p, d, q). For financial price data, it typically selects **d = 1** (one differencing step) since prices are non-stationary.
> - The estimated coefficients represent the temporal dependency structure. An ARIMA(0,1,0) is equivalent to a **random walk** — the standard benchmark in finance.
> - The reported **AIC** and **BIC** are used to compare alternative specifications: lower values indicate better complexity-penalized fit.

## Parameter Significance

```{r arima-coeftest}
cat("=== ARIMA - Parameter Significance ===\n")
coeftest(fit_arima)
```

> **Interpretation:** The Wald test evaluates whether each parameter is statistically different from zero (H0: coefficient = 0). A p-value < 0.05 indicates significance at the 5% level. If the selected model is a random walk, no coefficients will appear in this test — which is expected and valid for financial series.

## Residual Diagnostics

```{r arima-residuals, fig.cap="ARIMA residual diagnostics: residuals should behave as white noise (no autocorrelation, approximately normal distribution)."}
checkresiduals(fit_arima)
```

> **Interpretation:**
>
> - **Top panel (residuals over time):** Should show no trend or changing variance. Volatility clustering is common and expected in financial data.
> - **ACF of residuals:** No bar should exceed the confidence bands (+/-1.96/sqrt(n)). Significant residual autocorrelation indicates model misspecification.
> - **Histogram:** Residuals should approximate a zero-centered normal distribution. Heavy tails are common in financial returns.
> - **Ljung-Box test:** A p-value > 0.05 means we fail to reject residual independence — the model adequately captures the temporal structure.

## ARIMA Forecast

```{r arima-forecast}
pred_arima <- forecast(fit_arima, h = h)
```

```{r arima-forecast-plot, fig.cap="ARIMA forecast with 80% and 95% confidence intervals. The orange line shows actual values from the test set."}
# Build all components as data frames with real dates
test_dates <- index(test)

# Full training history for context (series is now 2020-2025, so show all of it)
train_tail_df <- data.frame(
  date  = index(train),
  price = as.numeric(train)
)

# Forecast mean + confidence bands
forecast_df_arima <- data.frame(
  date  = test_dates,
  mean  = as.numeric(pred_arima$mean),
  lo80  = as.numeric(pred_arima$lower[, 1]),
  hi80  = as.numeric(pred_arima$upper[, 1]),
  lo95  = as.numeric(pred_arima$lower[, 2]),
  hi95  = as.numeric(pred_arima$upper[, 2])
)

# Actual test values
test_df <- data.frame(
  date  = test_dates,
  price = as.numeric(test)
)

ggplot() +
  # 95% CI band
  geom_ribbon(data = forecast_df_arima,
              aes(x = date, ymin = lo95, ymax = hi95),
              fill = "steelblue", alpha = 0.15) +
  # 80% CI band
  geom_ribbon(data = forecast_df_arima,
              aes(x = date, ymin = lo80, ymax = hi80),
              fill = "steelblue", alpha = 0.25) +
  # Training history
  geom_line(data = train_tail_df,
            aes(x = date, y = price, colour = "Training"),
            linewidth = 0.5) +
  # Forecast mean
  geom_line(data = forecast_df_arima,
            aes(x = date, y = mean, colour = "Forecast"),
            linewidth = 0.8, linetype = "dashed") +
  # Actual test values
  geom_line(data = test_df,
            aes(x = date, y = price, colour = "Actual"),
            linewidth = 0.8) +
  scale_colour_manual(
    values = c("Training" = "black", "Forecast" = "steelblue", "Actual" = "darkorange"),
    breaks = c("Training", "Forecast", "Actual")
  ) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title    = "ARIMA Forecast vs Actual Values",
       subtitle = paste0("Horizon: ", h, " trading days | Shaded: 80% and 95% CI"),
       y        = "Price (USD)", x = "Date", colour = "") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```

> **Visual interpretation:** ARIMA forecasts for financial series tend to project the **last observed trend** or a relatively flat level if the selected model is a random walk. Confidence intervals widen significantly over the horizon, reflecting the high inherent uncertainty of long-range forecasting for volatile assets.

---

# Model 2: ETS (Exponential Smoothing)

## Estimation

```{r ets-fit, cache=TRUE}
cat("=== ETS Fitting ===\n")
fit_ets <- ets(train)
summary(fit_ets)
```

> **Interpretation of ETS model:**
>
> The ETS model is specified through three components: **Error (E)**, **Trend (T)**, **Seasonality (S)**. Each can be:
> - **Additive (A):** components sum together
> - **Multiplicative (M):** components multiply together
> - **None (N):** component is absent
>
> For financial prices with a strong trend, `ets()` typically selects **ETS(M,A,N)** or **ETS(A,A,N)** — error and trend with no seasonality. The smoothing parameters alpha (level), beta (trend), and gamma (seasonal) control how quickly the model adapts to recent observations.

## Residual Diagnostics

```{r ets-residuals, fig.cap="ETS residual diagnostics."}
checkresiduals(fit_ets)
```

> **Interpretation:** As with ARIMA, we verify whether ETS residuals behave as white noise. Significant residual autocorrelation would indicate that the model is not fully capturing the temporal structure present in the data.

## ETS Forecast

```{r ets-forecast}
pred_ets <- forecast(fit_ets, h = h)
```

```{r ets-forecast-plot, fig.cap="ETS forecast with confidence intervals. ETS typically projects a linear trend extrapolated from the slope estimated at the end of the training period."}
forecast_df_ets <- data.frame(
  date = test_dates,
  mean = as.numeric(pred_ets$mean),
  lo80 = pmax(0, as.numeric(pred_ets$lower[, 1])),
  hi80 = as.numeric(pred_ets$upper[, 1]),
  lo95 = pmax(0, as.numeric(pred_ets$lower[, 2])),
  hi95 = as.numeric(pred_ets$upper[, 2])
)

ggplot() +
  geom_ribbon(data = forecast_df_ets,
              aes(x = date, ymin = lo95, ymax = hi95),
              fill = "tomato", alpha = 0.15) +
  geom_ribbon(data = forecast_df_ets,
              aes(x = date, ymin = lo80, ymax = hi80),
              fill = "tomato", alpha = 0.25) +
  geom_line(data = train_tail_df,
            aes(x = date, y = price, colour = "Training"),
            linewidth = 0.5) +
  geom_line(data = forecast_df_ets,
            aes(x = date, y = mean, colour = "Forecast"),
            linewidth = 0.8, linetype = "dashed") +
  geom_line(data = test_df,
            aes(x = date, y = price, colour = "Actual"),
            linewidth = 0.8) +
  scale_colour_manual(
    values = c("Training" = "black", "Forecast" = "tomato", "Actual" = "darkorange"),
    breaks = c("Training", "Forecast", "Actual")
  ) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title    = "ETS Forecast vs Actual Values",
       subtitle = paste0("Horizon: ", h, " trading days | Shaded: 80% and 95% CI"),
       y        = "Price (USD)", x = "Date", colour = "") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```

> **Visual interpretation:** ETS projects a **smoothed trend**. If the model captured the upward trend from the training period, it will forecast rising prices. The projected growth rate depends on the beta parameter. Given NVDA's near-exponential behavior in 2022-2025, even trend-aware models may underestimate the actual growth.

---

# Model 3: Holt-Winters

```{r hw-setup}
# Convert to ts object with frequency 5 (5-day trading week)
train_ts <- ts(as.numeric(train), frequency = 5)
```

## Holt-Winters Additive

```{r hw-additive, cache=TRUE}
cat("=== Holt-Winters Additive ===\n")
fit_hw_add <- hw(train_ts, h = h, seasonal = "additive")
summary(fit_hw_add)
```

> **Interpretation (Additive):** The additive model assumes seasonal fluctuations are **constant in absolute magnitude** over time (e.g., always +/-$5 on certain days of the week). The estimated parameters are:
> - **alpha:** level smoothing — values near 1 place more weight on recent observations
> - **beta:** trend smoothing — controls how quickly the estimated slope changes
> - **gamma:** seasonal smoothing — controls adaptability of the seasonal component

## Holt-Winters Multiplicative

```{r hw-multiplicative, cache=TRUE}
cat("=== Holt-Winters Multiplicative ===\n")
fit_hw_mul <- hw(train_ts, h = h, seasonal = "multiplicative")
summary(fit_hw_mul)
```

> **Interpretation (Multiplicative):** The multiplicative model assumes seasonal fluctuations are **proportional to the series level** (e.g., +/-5% on certain days). For trending series like NVDA, the multiplicative model is generally more appropriate because absolute fluctuations grow alongside the price level. Parameters carry the same interpretation as in the additive variant.

## Optimal Holt-Winters Parameters

```{r hw-params}
cat("=== Optimal Parameters - HW Additive ===\n")
cat("Alpha (level)   :", fit_hw_add$model$par["alpha"], "\n")
cat("Beta  (trend)   :", fit_hw_add$model$par["beta"],  "\n")
cat("Gamma (seasonal):", fit_hw_add$model$par["gamma"], "\n")

cat("\n=== Optimal Parameters - HW Multiplicative ===\n")
cat("Alpha (level)   :", fit_hw_mul$model$par["alpha"], "\n")
cat("Beta  (trend)   :", fit_hw_mul$model$par["beta"],  "\n")
cat("Gamma (seasonal):", fit_hw_mul$model$par["gamma"], "\n")
```

> **Comparative interpretation:** The values of alpha, beta, and gamma reveal the model's "memory." An alpha = 0.99 means nearly all information comes from the most recent observation (highly reactive model). A beta close to 0 indicates that the trend changes slowly. Compare parameters between the additive and multiplicative models to understand their structural differences.

---

# Model Evaluation

## Metric Calculation

```{r metrics}
test_vec <- as.numeric(test)

rmse_arima  <- rmse(test_vec, as.numeric(pred_arima$mean))
mae_arima   <- mae(test_vec,  as.numeric(pred_arima$mean))
mape_arima  <- mape(test_vec, as.numeric(pred_arima$mean)) * 100
r2_arima    <- 1 - sum((test_vec - as.numeric(pred_arima$mean))^2) / sum((test_vec - mean(test_vec))^2)

rmse_ets    <- rmse(test_vec, as.numeric(pred_ets$mean))
mae_ets     <- mae(test_vec,  as.numeric(pred_ets$mean))
mape_ets    <- mape(test_vec, as.numeric(pred_ets$mean)) * 100
r2_ets      <- 1 - sum((test_vec - as.numeric(pred_ets$mean))^2) / sum((test_vec - mean(test_vec))^2)

rmse_hw_add <- rmse(test_vec, as.numeric(fit_hw_add$mean))
mae_hw_add  <- mae(test_vec,  as.numeric(fit_hw_add$mean))
mape_hw_add <- mape(test_vec, as.numeric(fit_hw_add$mean)) * 100
r2_hw_add   <- 1 - sum((test_vec - as.numeric(fit_hw_add$mean))^2) / sum((test_vec - mean(test_vec))^2)

rmse_hw_mul <- rmse(test_vec, as.numeric(fit_hw_mul$mean))
mae_hw_mul  <- mae(test_vec,  as.numeric(fit_hw_mul$mean))
mape_hw_mul <- mape(test_vec, as.numeric(fit_hw_mul$mean)) * 100
r2_hw_mul   <- 1 - sum((test_vec - as.numeric(fit_hw_mul$mean))^2) / sum((test_vec - mean(test_vec))^2)
```

## Model Comparison Table

```{r results-table}
results <- data.frame(
  Model = c("ARIMA", "ETS", "HW-Additive", "HW-Multiplicative"),
  RMSE  = round(c(rmse_arima, rmse_ets, rmse_hw_add, rmse_hw_mul), 4),
  MAE   = round(c(mae_arima,  mae_ets,  mae_hw_add,  mae_hw_mul),  4),
  MAPE  = round(c(mape_arima, mape_ets, mape_hw_add, mape_hw_mul), 4),
  R2    = round(c(r2_arima,   r2_ets,   r2_hw_add,   r2_hw_mul),   4)
)

results <- results %>%
  mutate(
    Rank_RMSE = rank(RMSE),
    Rank_MAE  = rank(MAE),
    Avg_Rank  = (Rank_RMSE + Rank_MAE) / 2
  )

best_rmse    <- results$Model[which.min(results$RMSE)]
best_mae     <- results$Model[which.min(results$MAE)]
best_overall <- results$Model[which.min(results$Avg_Rank)]

results %>%
  select(Model, RMSE, MAE, MAPE, R2, Rank_RMSE, Rank_MAE) %>%
  kable(
    caption   = "Model Comparison - RMSE, MAE, MAPE and R² on Test Set",
    col.names = c("Model", "RMSE (USD)", "MAE (USD)", "MAPE (%)", "R²", "RMSE Rank", "MAE Rank"),
    align     = c("l", "r", "r", "r", "r", "c", "c")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width        = FALSE,
    position          = "center"
  ) %>%
  row_spec(which.min(results$RMSE), bold = TRUE, color = "white",
           background = "#2ecc71") %>%
  add_header_above(c(" " = 1, "Error Metrics" = 4, "Rankings" = 2)) %>%
  footnote(general = paste0("Best model by RMSE: ", best_rmse,
                             " | Best model by MAE: ", best_mae))
```

> **Metric interpretation:**
>
> - **RMSE (Root Mean Squared Error):** Penalizes large errors more heavily due to squaring. Sensitive to outliers. Expressed in the same unit as price (USD). An RMSE of $20 means the model's squared-average error corresponds to roughly $20 per day.
> - **MAE (Mean Absolute Error):** Average of absolute errors. More robust to outliers. Represents the typical expected error in dollars.
> - **Preference:** If large errors are disproportionately costly (e.g., in investment decisions), minimizing RMSE is preferred. For general evaluation, MAE is more directly interpretable.

## RMSE Comparison Chart

```{r plot-rmse, fig.height=5, fig.cap="RMSE comparison across models. Shorter bar = better model. The dashed line marks the overall average."}
ggplot(results, aes(x = reorder(Model, RMSE), y = RMSE, fill = Model)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = paste0("$", round(RMSE, 2))),
            vjust = -0.5, fontface = "bold", size = 4.5) +
  geom_hline(yintercept = mean(results$RMSE),
             linetype = "dashed", color = "gray40", linewidth = 0.8) +
  annotate("text",
           x = 0.6, y = mean(results$RMSE) * 1.03,
           label = "Average", color = "gray40", size = 3.5) +
  scale_fill_manual(values = c("ARIMA"             = "steelblue",
                                "ETS"               = "tomato",
                                "HW-Additive"       = "seagreen",
                                "HW-Multiplicative" = "goldenrod")) +
  labs(title    = "Model Comparison - RMSE",
       subtitle = "Root Mean Squared Error (USD) | Test Set (25%)",
       x        = "Model",
       y        = "RMSE (USD)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## MAE Comparison Chart

```{r plot-mae, fig.height=5, fig.cap="MAE comparison across models. MAE represents the average absolute prediction error in USD."}
ggplot(results, aes(x = reorder(Model, MAE), y = MAE, fill = Model)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = paste0("$", round(MAE, 2))),
            vjust = -0.5, fontface = "bold", size = 4.5) +
  geom_hline(yintercept = mean(results$MAE),
             linetype = "dashed", color = "gray40", linewidth = 0.8) +
  annotate("text",
           x = 0.6, y = mean(results$MAE) * 1.03,
           label = "Average", color = "gray40", size = 3.5) +
  scale_fill_manual(values = c("ARIMA"             = "steelblue",
                                "ETS"               = "tomato",
                                "HW-Additive"       = "seagreen",
                                "HW-Multiplicative" = "goldenrod")) +
  labs(title    = "Model Comparison - MAE",
       subtitle = "Mean Absolute Error (USD) | Test Set (25%)",
       x        = "Model",
       y        = "MAE (USD)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## Combined RMSE and MAE Chart

```{r plot-combined, fig.height=6, fig.cap="Simultaneous comparison of RMSE and MAE. The optimal model minimizes both metrics."}
results_long <- melt(results[, c("Model", "RMSE", "MAE")],
                     id.vars       = "Model",
                     variable.name = "Metric",
                     value.name    = "Value")

ggplot(results_long, aes(x = reorder(Model, Value), y = Value,
                          fill = Metric, group = Metric)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = round(Value, 1)),
            position = position_dodge(width = 0.7),
            vjust = -0.4, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("RMSE" = "steelblue", "MAE" = "darkorange"),
                    labels  = c("RMSE (penalizes large errors)",
                                "MAE  (average absolute error)")) +
  labs(title    = "Model Comparison - RMSE and MAE",
       subtitle = "Test Set | Lower value = better model",
       x        = "Model",
       y        = "Error (USD)",
       fill     = "Metric") +
  theme_minimal(base_size = 13) +
  theme(plot.title      = element_text(face = "bold"),
        legend.position = "bottom")
```


> **Visual interpretation:** This chart is key to understanding how closely each model tracks NVDA's actual trajectory during the test period. No classical linear model can perfectly capture the exponential behavior of the AI boom. This is reflected in elevated RMSE/MAE values across all models.

---

# Best Model Selection

```{r best-model}
cat("=========================================================\n")
cat("          FINAL SUMMARY - BEST MODEL                    \n")
cat("=========================================================\n")
cat(sprintf("  %-20s  %10s    %10s\n", "Model", "RMSE", "MAE"))
cat("---------------------------------------------------------\n")
for (i in 1:nrow(results)) {
  cat(sprintf("  %-20s  %10.4f    %10.4f\n",
              results$Model[i], results$RMSE[i], results$MAE[i]))
}
cat("---------------------------------------------------------\n")
cat("  Best by RMSE   :", best_rmse, "\n")
cat("  Best by MAE    :", best_mae,  "\n")
cat("  Best overall   :", best_overall, "\n")
cat("=========================================================\n")
```

```{r ranking-plot, fig.height=4, fig.cap="Aggregate model ranking (average of RMSE and MAE ranks). Lower value = better overall performance."}
ggplot(results, aes(x = reorder(Model, Avg_Rank), y = Avg_Rank, fill = Model)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = paste0("Rank: ", round(Avg_Rank, 1))),
            vjust = -0.5, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = c("ARIMA"             = "steelblue",
                                "ETS"               = "tomato",
                                "HW-Additive"       = "seagreen",
                                "HW-Multiplicative" = "goldenrod")) +
  labs(title    = "Overall Model Ranking (Average of RMSE + MAE Ranks)",
       subtitle = "Rank 1 = best model",
       x        = "Model",
       y        = "Average Rank") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

---

# Conclusions

## Results Summary

```{r conclusions-table}
summary_tbl <- data.frame(
  Aspect = c(
    "Time series",
    "Stationarity",
    "Applied transformation",
    "Best model (RMSE)",
    "Best model (MAE)",
    "Best overall model"
  ),
  Finding = c(
    "Non-stationary (strong exponential trend)",
    "KPSS rejects H0 on original series; log-returns are stationary",
    "First difference of the logarithm (log-returns)",
    best_rmse,
    best_mae,
    best_overall
  )
)

kable(summary_tbl, col.names = c("Aspect", "Finding"),
      caption = "Executive Analysis Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = TRUE) %>%
  column_spec(1, bold = TRUE, width = "30%")
```

## Final Discussion

**1. On the difficulty of financial forecasting:**  
NVDA's price exhibited **exceptional behavior** during the test period, driven by macroeconomic and technological factors that no univariate statistical model can anticipate. This explains the relatively high errors observed across all models.

**2. Model-by-model behavior:**  
- **ARIMA:** By modeling differences, it produces conservative forecasts close to the last observed value. It is the most theoretically rigorous approach for financial series.  
- **ETS:** Captures the smoothed trend but may over- or under-forecast when the trend changes abruptly.  
- **HW Additive vs Multiplicative:** For series with a growing trend, the multiplicative model generally captures proportional dynamics better. The artificial 5-day frequency limits the usefulness of the seasonal component.

**3. Limitations:**  
- Classical models assume **linearity and homoscedasticity** — assumptions violated by NVDA's extreme behavior.  
- For financial data, models such as **GARCH** (captures time-varying volatility) or **machine learning models** (LSTM, XGBoost) could yield better forecasting accuracy.  
- The forecast horizon (~1,600 days) is very long for any time series model; errors accumulate substantially over time.

**4. Recommendation:**  
The **`r best_overall`** model performs best under the evaluated metrics. However, for real-world investment decisions, it is strongly recommended to complement this analysis with fundamental analysis, stochastic volatility models, and ensemble forecasting methods.


*Report automatically generated with R Markdown | IS512 - Time Series Analysis*
